---
title: "forecasting"
format: 
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
editor: visual
execute:
  echo: true
  warning: false
---

```{r}
library(dplyr) 
library(lubridate) 
library(xts) 
library(tsoutliers) 
library(parallel) 
library(doParallel) 
library(seastests)
library(tseries)
```

Let's clean the data:

```{r}
df<- read.csv("Data/sacrois_complete.csv")
print(unique(df$X3A_CODE))
print(length(unique(df$X3A_CODE))) # 44 species
df$DATE_SEQ <- as.Date(df$DATE_SEQ)
str(df)

df <- df %>%
  rename("Date" = "DATE_SEQ") %>%
  rename("Species" = "X3A_CODE") %>%
  rename("LandingsKG" = "QUANT_POIDS_VIF_MOYENNE") %>%
  rename("LandingsEUR" = "MONTANT_EUROS_MOYENNE")%>%
  filter(Date >= "2001-01-01" & Date < "2024-01-01")
```

We are getting rid of the late 1999 data and beginning of 2024 data.

# Create time series for each species

We will create a `ts_list()` , a list of lists, that contains the time-series of each species of the data set. The frequency is **365** because we are dealing with daily data.

```{r}
# Get the list of unique species
species_list <- unique(df$Species)

# Define the exact date range we want
start_date <- as.Date("2000-01-01")
end_date <- as.Date("2023-12-31")

# Create a complete date sequence
date_seq <- seq(start_date, end_date, by = "day")

# Create a time series for each species
ts_list <- list()
for (species in species_list) {
  # Filter data for the current species and aggregate by date
  species_data <- df %>% 
    filter(Species == species) %>% 
    group_by(Date) %>%
    summarize(LandingsKG = sum(LandingsKG, na.rm = TRUE)) %>%
    ungroup()
  
  # Create a data frame with all dates
  full_data <- data.frame(Date = date_seq)
  
  # Merge with the species data
  full_data <- left_join(full_data, species_data, by = "Date")
  
  # Fill NA values with 0
  full_data$LandingsKG[is.na(full_data$LandingsKG)] <- 0
  
  # Create a ts object
  ts_object <- ts(full_data$LandingsKG, 
                  start = c(2000, 1), 
                  frequency = 365)
  
  # Add the ts object to the list
  ts_list[[paste0("ts_", species)]] <- ts_object
}

# Create a vector with all the TS names
ts_names <- names(ts_list)

# Print the names of the time series
print(ts_names)
```

# Seasonality Detection

We want to apply the `combined_test` on each time series to sea if there is any seasonality. You should note that:

-   **H1** means **seasonality** *(alternative hypothesis)*

-   **H0** means **NO seasonality** *(null hypothesis)*

-   if p \< 0.05 then H1 : seasonality

```{r}
seasonal_combined_test <- c()
seasonal_seasdum <- c()

# Loop over each TS
for (ts_name in names(ts_list)) {
  ts_data <- ts_list[[ts_name]]  # Retrieve data from the time series list
  
  # Apply the combined_test() test to the time series
  ct_res <- combined_test(ts_data)
  
  # Check each p-value and display results if at least one valid H1
  ct_results <- ct_res$stat
  ct_pvals <- ct_res$Pval
  if (ct_results == TRUE) {
    cat("\n")
    cat("\n")
    print(paste0("Combined_test results for the series ", ts_name))
    print(ct_res)
    seasonal_combined_test <- c(seasonal_combined_test, ts_name)
  }
  
  # # Apply the seasdum() test to the time series
  # sd_res <- seasdum(ts_data)
  # 
  # # Check p-value and display results if < 0.05
  # if (sd_res$Pval < 0.05) {
  #   cat("\n")
  #   cat("\n")
  #   print(paste0("Résultats du seasdum pour la série ", ts_name))
  #   print(sd_res)
  #   seasonal_seasdum <- c(seasonal_seasdum, ts_name)
  # }
}

assign("seasonal_combined_test", seasonal_combined_test, envir = .GlobalEnv)
```

[The `combined_test()` function combines three tests:]{.underline}

-   QS test (Quantile Spectral test)

-   QS-R test (QS test on residuals)

-   KW-R test (Kruskal-Wallis test on residuals)

[The function considers a time series as seasonal if:]{.underline}

-   The p-value of QS-R is \< 0.01 OR the p-value of KW-R is \< 0.001

-   OR if both the p-value of QS is \< 0.01 AND the p-value of KW-R is \< 0.01

[Here are the time series that have Seasonality:]{.underline}

```{r}
seasonal_combined_test
```

We need to correct that seasonality among those time series.

```{r}
for (ts_name in seasonal_combined_test) {
  ts_data <- ts_list[[ts_name]]  # Retrieve time series data from the list
  decomp <- stl(ts_data, s.window = 365)  # STL decomposition
  seasonal <- decomp$time.series[, "seasonal"]  # Get the seasonal component
  ts_data_adjusted <- ts_data - seasonal  # Correcting the seasonal component
  ts_list[[ts_name]] <- ts_data_adjusted  # Update the adjusted time series in the list
}
```

Now that this is done, let's check again for any seasonality:

```{r}
seasonal_combined_test <- c()
seasonal_seasdum <- c()

# Loop over each TS
for (ts_name in names(ts_list)) {
  ts_data <- ts_list[[ts_name]]  # Retrieve data from the time series list
  
  # Apply the combined_test() test to the time series
  ct_res <- combined_test(ts_data)
  
  # Check each p-value and display results if at least one valid H1
  ct_results <- ct_res$stat
  ct_pvals <- ct_res$Pval
  if (ct_results == TRUE) {
    cat("\n")
    cat("\n")
    print(paste0("Combined_test results for the series ", ts_name))
    print(ct_res)
    seasonal_combined_test <- c(seasonal_combined_test, ts_name)
  }
}
```

It's weird that we have such a difference between the first two tests and the last one. There are quite the opposite, in term of P-values.

For now we will focus on the first two tests *(QS & QS-R)* tests as our results for seasonality. According to them, **there is no seasonality left.**

# Stationarity

Now, let's take care of stationarity. Most of the forecasting models we are going to use need stationary time-series. We will use the `adf.test` from the [tseries](https://www.rdocumentation.org/packages/tseries/versions/0.10-54/topics/adf.test) package.

Note that for this test we have:

-   H0: série non stationnaire (possède une racine unitaire)

-   H1: série stationnaire (n'en possède pas)

-   if p\<0.05 : the serie is stationary

```{r}
#| warning: false
# First round of differencing
for (ts_name in names(ts_list)) {
  ts_data <- ts_list[[ts_name]]  # Retrieve time series data from the list
  adf_result <- adf.test(ts_data)  # Apply ADF test for stationarity
  #print(adf_result)
  
  # Check if P-Value > 0.05
  if (adf_result$p.value > 0.05) {
    # Apply a difference to the time series if it is not stationary
    ts_list[[ts_name]] <- diff(ts_data)
    print(paste("The TS", ts_name, "has been differentiated"))
  }
  else {
    print(paste("The TS", ts_name, "is stationary"))
  }
}

# Second round of differencing if necessary
for (ts_name in names(ts_list)) {
  ts_data <- ts_list[[ts_name]]  # Retrieve time series data from the list
  adf_result <- adf.test(ts_data)  # Apply ADF test for stationarity
  #print(adf_result)
  
  # Check if P-Value > 0.05
  if (adf_result$p.value > 0.05) {
    print(paste("The TS", ts_name, "is still not stationary"))
    # Apply a difference to the time series if it is not stationary
    ts_list[[ts_name]] <- diff(ts_data)
    print(paste("The TS", ts_name, "has been differentiated a second time"))
  }
}

# Final check
for (ts_name in names(ts_list)) {
  ts_data <- ts_list[[ts_name]]  # Retrieve time series data from the list
  adf_result <- adf.test(ts_data)  # Apply ADF test for stationarity
  #print(adf_result)
  
  if (adf_result$p.value > 0.05) {
    # Check if P-Value > 0.05
    cat(paste0("\033[1m\033[4m\033[31mThe time series ", ts_name, " is still not stationary:\033[0m\n"))
  }
}
```

It seems that every time-series is now stationary. Let's check if they all have the same length

```{r}
str(ts_list)
```

They do. Now, before forecasting we are going to use a variables selection method. This will help us to reduce the number of irrelevant variables of our models.

After that, we will apply econometrics and machine learning forecasting models.

From now on, we'll reason as follows:

-   $Y_t$: will be the variable we seek to predict

-   $X_k,_t$: will be the explanatory variables, which will add information to enable the models to better predict our $Y_t$.

For your information, here's a table summarizing our variables. In bold are the dynamic variables in [Ifremer's IAM forecasting model](https://archimer.ifremer.fr/doc/00784/89579/). As these variables are already considered dynamic, they are the only ones we won't attempt to predict.

| HKE     | European hake              | MUT     |                          |
|---------|----------------------------|---------|--------------------------|
| **NEP** |                            | **ARA** |                          |
| **DPS** |                            | ANE     | European anchovy         |
| BES     | Belone                     | BOG     | Bogue                    |
| BSS     | European seabass           | BZX     | Bonitos                  |
| CTL     | Cuttlefish, bobtail squids | DCP     | Natantian decapods       |
| DEX     | Dentex                     | DPS     | Deep-water rose shrimp   |
| ELX     | River eels                 | FIN     | Finfishes                |
| FLX     | Flatfishes                 | FOX     | Forkbeards               |
| GUX     | Gurnards, searobins        | JAX     | Jack and horse mackerels |
| JLX     | Murex shells               | JOD     | John dory                |
| MAX     | Mackerels                  | MGR     | Meagre                   |
| MNZ     | Monkfishes                 | MUL     | Mullets                  |
| MUX_bis | Red mullets                | OCT     | Octopuses, etc.          |
| PAC     | Common pandora             | PEN     | Penaeus shrimps          |
| PIL     | European pilchard          | POD     | Poor cod                 |
| SAA     | Round sardinella           | SBA     | Axillary seabream        |
| SBG     | Gilthead seabream          | SBX     | Porgies, seabreams       |
| SOX     | Soles                      | SQY     | Squillids                |
| SQZ     | Inshore squids             | SRX     | Rays, stingrays, mantas  |
| SWM     | Swimming crabs, etc.       | TUX     | Tuna-like fishes         |
| VLO     | Spiny lobsters             | WEX     | Weevers                  |
| WHB     | Blue whiting               | XOX     | Sandlances               |

: Different species of the SACROIS dataset

[Note]{.underline}: `ZZZ` therefore includes all species not mentioned in the table above. It is therefore a collection of several species.
